A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-01-28 paper.tex
worked on the intro

Reactive systems are ubiquitous in real-world problems such as circuit design,
industrial automation, or device drivers. Automatic synthesis can provide a
\emph{correct by construction} controller for a reactive system from a
specification.  However, the reactive synthesis problem is 2EXPTIME-complete so
naive algorithms are infeasible on anything but simple systems.

Reactive synthesis is formalised as a game between the \emph{controller} and
its \emph{environment}. In this work we focus on safety games, in which the
controller must prevent the environment from forcing the game into an error
state. In this way the environment is equivalent to an existential search of
the game and the controller is universal. Much of the complexity of reactive
synthesis stems from managing the interactions of the alternating quantifiers
on the state space of the game.

There are several techniques that aim to mitigate this complexity by
representing states symbolically.  Historically the most successful technique
has been to use \emph{Binary Decision Diagrams}.  BDDs efficiently
represent a relation on a set of game variables but in the worst case the
representation may be exponential in the number of variables. As a result
BDDs are not a one-size-fits-all solution for all reactive synthesis
specifications.

Advances in SAT solving technology has prompted research into its applicability
to synthesis as an alternative to BDDs. One approach is to find sets of states
in CNF \cite{bloem2014,morgenstern2013}. Another approach is to eschew states
and focus on \emph{runs} of the game. Previous work has applied this idea to
realizability of bounded games \cite{narodytska2014} by forming abstractions of
the game and refining in a counterexample-guided framework. This has been shown
to outperform BDDs on certain classes of game specifications but is only able
to find counterexamples to the safety property of a certain length.  In this
paper, we extend this idea to unbounded games by approximating sets of unsafe
states from abstract games. Careful construction ensures that a fixed point in
these sets guarantees completeness.

@@

    \begin{algorithmic}
        \algrestore{b1}
        \Function{findCandidate}{$p, s, k, T$}
            \State $f \gets $ \IfElse{$p = \texttt{cont}$}{\Call{treeFormula}{$T$}}{\Call{\textoverline{treeFormula}}{$T$}} \EndIfElse
            \State $sol \gets $ \Call{SAT}{$s(X_T) \land f$}
            \If{$sol = \texttt{unsat}$} 
                \If{\texttt{unbounded}} \Comment{Active only in the unbounded solver}
                    \State $\sigma \gets $ \Call{generalise}{$s$} \Comment{Expand $s$ to a set of states}
                    \State \IfElse{$p = \texttt{cont}$}{\Call{learn}{$\sigma, T$}}{\Call{\textoverline{learn}}{$\sigma, T$}} \EndIfElse
                \EndIf
                \State \Return $\emptyset$ \Comment{No candidate exists}
            \Else
                \State \Return $\{ \langle n, c \rangle | n \in $ \Call{gtNodes}{$T$} $, c = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
            \EndIf
        \EndFunction
        \algstore{b2}
    \end{algorithmic}

>>>>

Reactive systems are ubiquitous in real-world problems such as circuit design,
industrial automation, or device drivers. Automatic synthesis can provide a
\emph{correct by construction} controller for a reactive system from a
specification.  Reactive synthesis is formalised as a game between the \emph{controller} and
its \emph{environment}. In this work we focus on safety games, in which the
controller must prevent the environment from forcing the game into an error
state.

The reactive synthesis problem is 2EXPTIME-complete so
naive algorithms are infeasible on anything but simple systems.
% In this way the environment is equivalent to an existential search of
%the game and the controller is universal. Much of the complexity of reactive
%synthesis stems from managing the interactions of the alternating quantifiers
%on the state space of the game.
There are several techniques that aim to mitigate this complexity by
representing states and transitions of the system symbolically~\cite{Piterman_PS_06,bloem2014,morgenstern2013}. 
These techniques incrementally construct a symbolic representation of the set of 
discovered winning states of the game.  The downside of this approach is that 
keeping track of all discovered winning states can lead to a space explosion even
when using efficient symbolic representation such as BDDs or CNFs.

%Historically the most successful technique
%has been to use \emph{Binary Decision Diagrams}.  BDDs efficiently
%represent a relation on a set of game variables but in the worst case the
%representation may be exponential in the number of variables. As a result
%BDDs are not a one-size-fits-all solution for all reactive synthesis
%specifications.

%Advances in SAT solving technology has prompted research into its applicability
%to synthesis as an alternative to BDDs. One approach is to find sets of states
%in CNF \cite{bloem2014,morgenstern2013}.

An alternative approach, proposed by Narodytska et al.~\cite{narodytska2014},
is to eschew states and focus on \emph{runs} of the game.  The method works by
exploring a subset of the concrete runs of the game and proving that these runs can be
generalised into a winning strategy on behalf of one of the players. In contrast to 
other existing synthesis methods, it does not store,
in either symbolic or explicit form, the set of winning states.
Instead, it uses counterexample-guided backtracking search to identify a small subset
of runs that are sufficient to solve the game.

This method has been shown to outperform BDDs on certain classes of games; however it
suffers from an important limitation: it is only able to solve games with a bounded number
of rounds.  In case of safety games, this means proving that the controller 
can keep the game within the safe region for a bounded number of steps.  
This is insufficient in most practical situations that require unbounded realizability.

%Previous work has applied this idea to
%realizability of bounded games \cite{narodytska2014} by forming abstractions of
%the game and refining in a counterexample-guided framework.

In this paper, we extend the method by Narodytska et al.~\cite{narodytska2014} to unbounded 
safety games.  To this end we enhance the method with computational learning: every time
the search algorithm discovers a new counterexample, the learning procedure analyzes the 
counterexample, extracting a subset of states winning for one of the players from it.  
The learning procedure ensures that reaching a fixed point in these sets is sufficient to establish 
unbounded realizability.
Our method can be seen as a hybrid between the counterexample-guided algorithm by
Narodytska et al. and methods based on winning set computation: we use the former to guide 
the search, while we rely on the latter to ensure convergence.

We evaluate our method on the benchmarks of the 2015 synthesis competition
(SYNTCOMP'15).  While our solver solves fewer total instances than competitors,
it solves the largest number of unique instances, i.e., instances that could not be solved by 
any other solver.  These results confirm that there exist classes of problems that are
hard for traditional synthesis techniques, but can be efficiently solved by our method.
While further performance improvements are clearly needed, in its current state the method 
can be used in a portfolio solver to increase the number of tractable instances.

@@

    \begin{algorithmic}
        \algrestore{b1}
        \Function{findCandidate}{$p, s, k, T$}
            \State $f \gets $ \IfElse{$p = \texttt{cont}$}{\Call{treeFormula}{$T$}}{\Call{\textoverline{treeFormula}}{$T$}} \EndIfElse
            \State $sol \gets $ \Call{SAT}{$s(X_T) \land f$}
            \If{$sol = \texttt{unsat}$} 
                \If{\texttt{unbounded}} \Comment{Active only in the unbounded solver}
                    \State $\sigma \gets $ \Call{generalise}{$s$} \Comment{Expand $s$ to a set of states}
                    \State \IfElse{$p = \texttt{cont}$}{\Call{learn}{$\sigma, T$}}{\Call{\textoverline{learn}}{$\sigma, T$}} \EndIfElse
                \EndIf
                \State \Return $\emptyset$ \Comment{No candidate exists}
            \Else
                \State \Return $\{ \langle n, c \rangle | n \in $ \Call{nodes}{$T$} $, c = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
            \EndIf
        \EndFunction
        \algstore{b2}
    \end{algorithmic}
